---
title: "R Notes"
author: "zwl"
date: "Long, long ago..."
output: html_document
---

# `R` Notes

## Data Types

`NA` represents missing data. 
`NULL`` represents that the value in quesiton does not exist.

**Vectors, matrices and arrays** must hold data of the **same** type. 
**data.frame** can have columns of different types, but the lengths are equal.
**Lists** can have different datatypes and can different sizes / dimensions.

Matrices are stored in **column major order**. To take input that is row by
row, use `matrix(..., byrow = T)`.

`List` - Indexing with `[]` returns list, a subset of the orignial. `[[]]` 
uses the data type of the element.

Standard (conditional) indexing includes `NA`. To exclude `NA`, use `subset()`.

To delete a component from a list: set the value to NULL. 

Functions to check data types:

* `mode()`
* `typeof()`
* `class()`

When added vectors of different lengths, `R` **recycles** the shorter vector,
i.e. repeats its value until the sizes are matched. I think this is so odd
and can be very dangerous... The saving grace is that the longer vector must
be integer multiples of the shorter one in length, i.e. 6 vs 2 works, but 5 vs 
2 does not.

`c()` will use the most common denominator of data types. It also will
**squeeze / flatten** the inputs.

```{r c}
a <- c(1, 2, 3)
b <- c(5, 6)
c <- c(a, c(8, 9), b)
c
```

Sequences: `:` operator generates **int**, `seq()` and `c()` generate 
**float**.

```{r identical}
a <-  1:3
b <- seq(1, 3, by=1)
mode(a)
mode(b)
identical(a, b)
```

`data.frame` has `strongAsFactors=` parameter to determine whether string 
columns should be factors.

`sapply()` sequeezes output of `lapply()` into a single vector or matrix. 
Both can take either list, matrices or data.frames.

`tapply()` groups values in an input vector by a list of factors, than apply
a function. `split()` does the grouping but does not apply a function.

`by()` works like `tapply()` but takes objects rather than vectors.

`table()` creates **contingency tables**. `subtable()` selects a subset.

`addmargins()` computes marginal values. 

`<<-` superassignment operator, or use `assign()` to change global variables.

On object memory usage, `R` doesn't have pointers... Instead when objects are
assigned to multiple variables, they reference the same memory location, 
i.e. the same copy, **until a modification is made**. 
At this point a copy is then made. 

For objects modified in an inner scope, the outer scope values are **not**
changed.

`pmin()` for pair-wise minimum.

Set operators:

* `union()`
* `intersect()`
* `setdiff()`
* `setequal()`
* `choose(n, k)`

**IO** functions:

* `scan()`
* `readline()`
* `cat()` print to screen
* Connection is R's version of a file handler, `file()`, `url()`, `seek()` and
`close()`, can open remote URLs like a file.
* `file.info()`, `dir()` to list files, `file.exists()`, `getwd()`, `setwd()`


**String** manipulation

`grep(pattern, x)`, `nchar(x)` to find length of `x`, `paste(a, b, sep=',')` 
to concatenate, `sprintf()` for formatting printing.

`substr(x, start, stop)`, `strsplit(x, split)`, `regexpr(pattern, text)`,
`grepexpr(pattern, text)` find all instances.

**Debug** tools

* `browser()` to set break in code, or `browser(condition)`
* `setBreakpoint(file, line_num)`
* `untrace()` to cancel
* `trace(f, t)` to call func `t` whenever func `f` is called
* call `traceback()` when errors occur

Useful global environmental vars to set:

```{r env, eval = F}
# dump when error occurrs, after a crash, call debugger() to examine
options(error=dump.frames)
# to alwyas enter debug mode:
# options(error=recover)
# to turn off
# options(error=NULL)

# show runtime errors instead of warnings
options(warn=2)
```

**Performance**

Use `system.time()` to time code.

`rm(list = ls())` to remove objects from memory, `gc()` to call garbage
collector.

Changing a value in a vector may cause the entire vector to be recomputed. 
Use `tracemem(x)` to find the memory location of `x`.

To profile code:

```{r rprof, eval = F}
Rprof()
# some code to profile
Rprof(NULL)
summaryRprof()
```

Use `invisible(code)` to hide result/output from code.

Some packages that handle large amount of data: `ff`, `bigmemory`.